---
title: "Report_HW3P3"
author: "Johnny Martinez"
date: "3/23/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Exercise J-2.2

## Helper functions to use for optim()
- Functions:
1. param_convert() - combines mu and sigma into one vector and can breakdown theta vector into mu and sigma
2. diff_xi_mu() - computes xi-mu and C := sum of (xi-mu)(xi-mu)^T 
3. log_like_mvn() - computes the log-likelihood for a multivariate normal
4. gradient() - computes the gradient

```{r}
param_convert <- function(mu=NULL, sigma=NULL, theta=NULL, t_comp=FALSE, ms_comp=FALSE){
    # decompose mu and sig into theta vector
    if(t_comp){t <- matrix(c(mu, sigma[upper.tri(sigma, TRUE)]),ncol=1)}
    # decompose theta vector into mu and sig 
    if(ms_comp){
        p <- sqrt(2*length(theta) + 9/4) - 3/2 # formula derived from num of thetas = p + p(p+1)/2
        mu <- matrix(theta[1:p],p,1) # create new mu vector
        
        sig <- matrix(0,p,p) # create new sig matrix
        sig[upper.tri(sig,TRUE)] <- theta[-c(1:p)]
        sig[lower.tri(sig)] <- t(sig)[lower.tri(sig)]
    }
    #writeLines("leave param_conv function")
    list(t=if(t_comp){t}, mu=if(ms_comp){mu}, sig=if(ms_comp){sig})
}

diff_xi_mu <- function(mu, comp_C=FALSE, comp_sxm=FALSE){
    # Compute C defined as sum of (xi-mu)(xi-mu)^T 
    xm <- apply(datan, 1, function(x) x-mu) # subtract mu from rows of x
    C <- xm %*% t(xm)        
    
    # Compute sum of (xi-mu) for use in gradient
    if(comp_sxm){
        sxm <- matrix(rowSums(xm), nrow=ncol(datan),1)
    }
    
    list(C = if(comp_C) C, sxm = if(comp_sxm) sxm)
}

log_like_mvn <- function(theta){
    n <- nrow(datan) # num of rows
    p <- ncol(datan) # num of columns
    
    t <- param_convert(theta=theta, ms_comp=TRUE)
    sigma <- t$sig
    mu <- t$mu
                       
    if(any(eigen(sigma)$values <= 0)){
        sigma[,] <- -Inf
    }
    
    C <- diff_xi_mu(mu, comp_C=TRUE)$C # C := sum of (xi-mu)(xi-mu)^T 
    
    log_det_sig <- log(det(sigma)) # log determinant of sigma
    sig_inv <- solve(sigma) # sigma inverse
    
    # Compute log likelihood function for multivariate normal
    # sum(sig.inv * C) = trace(sig.inv %*% C)
    log_like <- (-1/2)*(n*p*log(2*pi)+n*log_det_sig + sum(sig_inv * C ))
    log_like
    
}

gradient <- function(theta, t_comp=TRUE, dmu_comp=FALSE, dsig_comp=FALSE){    
    t <- param_convert(theta=theta, ms_comp=TRUE)
    mu <- t$mu
    sigma <- t$sig
    
    if(any(eigen(sigma)$values <= 0)){
        sigma[,] <- -Inf
    }
    
    #writeLines("\n entered in gradient function")
    sig_inv <- solve(sigma)
    diff <- diff_xi_mu(mu, comp_C=TRUE, comp_sxm=TRUE)
    sxm <- diff$sxm
    C <- diff$C
    
    # partial deriv_ formulas for mu indices
    dmu <- sig_inv %*% sxm
    
    # partial deriv_ formulas for sigma indices
    off_d_mat <- -(n * sig_inv - sig_inv %*% C %*% sig_inv) # take off diag elem_
    diag_mat <- (-1/2)*(n*sig_inv - sig_inv %*% C %*% sig_inv) # take diag elem_
    
    # dsig
    dsig <- matrix(0, nrow=p, ncol=p) # initialize pxp matrix for gradient of sigmas
    dsig[upper.tri(dsig)] <- off_d_mat[upper.tri(off_d_mat)] # fill matrix with lower tri of off_d_mat
    dsig[lower.tri(dsig)] <- t(dsig)[lower.tri(dsig)] # fill matrix with upper tri of off_d_mat
    diag(dsig) <- diag(diag_mat) # fill matrix with diag elements diagmat
    
    # theta form
    t <- param_convert(mu=dmu, sigma=dsig, t_comp=TRUE)$t
    t
}

```


## Provided function to generate data and print first 3 rows
```{r}

sqrtm <- function (A) {
    # Obtain matrix square root of a matrix A
    a = eigen(A)
    sqm = a$vectors %*% diag(sqrt(a$values)) %*% t(a$vectors)
    sqm = (sqm+t(sqm))/2
}

# Generate data

gen <- function(n,p,mu,sig,seed = 534){
    #---- Generate data from a p-variate normal with mean mu and covariance sigma
    # mu should be a p by 1 vector
    # sigma should be a positive definite p by matrix
    # Seed can be optionally set for the random number generator
    set.seed(seed)
    # generate data from normal mu sigma
    x = matrix(rnorm(n*p),n,p)
    datan = x %*% sqrtm(sig) + matrix(mu,n,p, byrow = TRUE)
    datan
}

# sample data
n <- 200
p <- 3
sigma <- matrix(c(1,.7,.7, .7, 1, .7, .7, .7, 1),p,p)
mu <- matrix(c(-1,1,2),p,1)
datan = gen(n,p,mu,sigma,seed = 2022)
```

### 3 first rows of data
```{r, echo=FALSE}
print(datan[1:3,])
```


## Use optim() with gradient function and log-likelihood function from above

```{r}
mu0 <- matrix(c(0,0,0), ncol=1)
sig0 <- diag(3)
theta0 <- param_convert(mu0, sig0, t_comp=TRUE)$t

optim(par=theta0, fn=log_like_mvn, gr=gradient, method="BFGS", 
      control = list(fnscale=-1, trace=100, abstol=1e-5))
```


# Exercise GH-2.3:

## Part (a) See attached hand-written page

## Part (b) 
```{r, message=FALSE}
library(formattable)
library(here)
data2 <- read.csv(here("data","censor_data.csv"))
```

```{r}
log_like_weib <- function(theta){
    a = theta[1]
    b0 = theta[2]
    b1 = theta[3]
    t = data2$t
    w = data2$w
    d = data2$d
    like <- sum((w*log(t^a * exp(b0+d*b1))) - t^a * exp(b0+d*b1) + w*log(a/t))
    like
}

gradient <- function(theta){
    a = theta[1]
    b0 = theta[2]
    b1 = theta[3]
    t = data2$t
    w = data2$w
    d = data2$d
    
    da <- sum(w/a + w*log(t) - t^a*exp(b0 + b1*d)*log(t))
    db0 <- sum(w - t^a*exp(b0 + b1*d))
    db1 <- sum(d*w - d*t^a*exp(b0 + b1*d))
    rbind(da,db0,db1)
}

hessian <- function(theta){
    a = theta[1]
    b0 = theta[2]
    b1 = theta[3]
    t = data2$t
    w = data2$w
    d = data2$d
    
    hess <- matrix(0, nrow=3, ncol=3)
    
    daa <- sum(-w/a^2 - t^a*exp(b0 + b1*d)*log(t)^2)
    dab0 <- sum(-t^a*exp(b0 + b1*d)*log(t))
    dab1 <- sum(-d*t^a*exp(b0 + b1*d)*log(t))
    fa <- cbind(daa,dab0,dab1)
    
    db0b0 <- sum(-t^a*exp(b0 + b1*d))
    db0b1 <- sum(-d*t^a*exp(b0 + b1*d))
    fb0 <- cbind(dab0, db0b0, db0b1)
    
    db1b1 <- sum(-d^2*t^a*exp(b0 + b1*d))
    fb1 <- cbind(dab1, db0b1, db1b1)
    
    hess[1,] <- fa
    hess[2,] <- fb0
    hess[3,] <- fb1
    
    hess
}

stop.criteria <- function(t0, t1, grad, tol.mre=1e-6, tol.grad=1e-5, itr, max.itr=100){
    mre <- max(abs((t1 - t0)/pmax(1, t1)))
    gradn <- norm(grad,"2")
    stop <- ifelse(itr == max.itr | gradn < tol.grad | mre < tol.mre, TRUE, FALSE)
    list(stop=stop, norm.g=gradn)
}

newton <- function(theta, tol.grad=1e-5, tol.mre=1e-6, max.itr=50){
    
    it <- 1
    stop <- FALSE
    
    while(!stop){
        obj.fn <- log_like_weib(theta) # set objective function        
        
        grad <- gradient(theta)
        hess <- hessian(theta)
        
        dir <- -solve(hess) %*% grad
        
        t1 <- theta + dir
        fun2 <- log_like_weib(t1)

        halve = 0
        while(fun2 < obj.fn & halve <= 20){
            halve = halve + 1
            t1 <- t0 + dir/2^halve
            fun2 <- log.like.weib(t1)
            print(c(it, halve))
        }
        if (halve >= 20) print('Step-halving failed after 20 halvings')
        
        obj.fn <- fun2
        
        stop.cri <- stop.criteria(theta, t1, grad=grad, itr=it)
        grad_norm <- stop.cri$norm.g
        
        
        stop <- stop.cri$stop
        
        it <- it + 1
        
        theta <- t1
        
        #print(c(it, halve,obj.fn, fun2))
        writeLines(paste("\nitr: ", it-1))
        writeLines(paste("halving   ", "log-like     ", "norm   "))
        writeLines(paste(halve, "        ", formattable(obj.fn, digits=4, format="f"), "   ", formattable(grad_norm, digits=1, format="e")))
        
        print ('-----------------------------------------')

    }
    theta
}

# initial value
theta <- c(1,0,0)
results <- newton(theta)
```

## Parameter estimation results using intial value = (1,1,1) 
```{r, echo=FALSE}
writeLines(paste0("alpha: ", round(results[1],7), "\nb0: ", round(results[2], 7), "\nb1: ", round(results[3],7)))
```

## Part (d)
### Compute standard errors using observed info matrix and then compute correlation matrix
```{r}
observed_info <- -solve(hessian(results))

sd_err <- sqrt(diag(observed_info))
```

```{r, echo=FALSE}
writeLines(paste0("std. errors: "))
sd_err
writeLines(paste0("correlation matrix: "))
cov2cor(V=observed_info)
```

* It can be seen in the correlation matrix that alpha is highly negatively correlated with b0,
while there is a slight correlation between alpha and b1. Moreover, there is almost
no correlation between b0 and b1. 